\documentclass[hyperref={bookmarks=false}]{beamer}

\usepackage[parfill]{parskip}
\usepackage{color}
\usepackage{listings}
\usepackage{hyperref}
\hypersetup{pdfauthor={Eugene Burmako},pdfsubject={Project Kepler},pdftitle={Project Kepler}}
\title{Project Kepler}

\definecolor{linkblue}{RGB}{49,57,174}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstdefinelanguage{scala}{
  morekeywords={abstract,annotation,case,catch,class,def,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,val,var,while,with,yield,
    macro},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/}
%  morestring=[b]",
%  morestring=[b]',
%  morestring=[b]"""
}

\lstset{frame=tb,
  language=scala,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  frame=single,
  breaklines=true,
  breakatwhitespace=true
  tabsize=3
}

\begin{document}

\title{Project Kepler}
\author{Eugene Burmako}
\institute{EPFL, LAMP}
\date{03 July 2012}
\maketitle

\begin{frame}[fragile]
\frametitle{Previously on Kepler}

\begin{itemize}
\item metaprogramming is useful
\item macros = compile-time AST transformers
\item prototype of macros for Scala
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Nowadays}

\begin{itemize}
\item there is a spec for macros
\item this spec is provisionally accepted for inclusion in Scala
\item starting from 2.10.0-M3 Scala has macros
\item the road to production has led to a number of discoveries
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Evolution}
v1:
\begin{lstlisting}[language=scala]
macro def assert(cond: Boolean, msg: Any) =
  if (assertionsEnabled)
    c"if (!$cond) raise(msg)"
  else
    c"()"
\end{lstlisting}

v2:
\begin{lstlisting}[language=scala]
def assert(cond: Boolean, msg: Any) = macro assertImpl
def assertImpl
    (c: Context)
    (cond: c.Expr[Boolean], msg: c.Expr[Any]) =
  if (assertionsEnabled)
    c.reify(if (!cond.splice) raise(msg.splice))
  else
    c.reify(())
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]
\frametitle{Reify}

The main insight we gained.

Kills two birds with one stone:
\begin{itemize}
\item implements a quasiquoting facility
\item achieves hygiene
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Hygiene}

Macro expansions are inlined into the call site.

What happens to bindings?

\begin{lstlisting}[language=scala]
object Assert {
  def raise(msg: Any) = throw new AssertionError(msg)
  macro def assert(cond: Boolean, msg: Any) =
    <[ if (!cond) raise(msg)) ]>
}

object Test extends App {
  def raise(msg: Any) = { /* haha, tricked you */ }
  assert(2 + 2 == 3, "no way")
}
\end{lstlisting}
\end{frame}

\begin{frame}[fragile]
\frametitle{Achieving hygiene}

In Common Lisp they have gensym.

In Scheme and Nemerle they perform transparent alpha-renaming.

In Scala we don't do any of this.
\end{frame}

\begin{frame}[fragile]
\frametitle{A fascinating discovery}

Non-hygienic macros can be bootstrapped into hygienic macros
(in that sense, our macros are self-cleaning).
\end{frame}

\begin{frame}[fragile]
\frametitle{Apparently that's old news}

\emph{Macros as Multi-Stage Computations} Ganz, Sabry \& Taha

\emph{Staged Notational Definitions} Taha \& Johann
\end{frame}

\begin{frame}[fragile]
\frametitle{Yet it's a novel interpretation}

Taha et al. build a macro system atop of a staged language.

We build a staged system atop of a macro language.
\end{frame}

\begin{frame}[fragile]
\frametitle{Deliverables}

\begin{itemize}
\item Paper: Scala Macros, a Technical Report (META'2012)
\item Documentation: \text{\color{linkblue}\href{http://scalamacros.org}{http://scalamacros.org}}
\item Code: \text{\color{linkblue}\href{http://github.com/scala/scala}{http://github.com/scala/scala}}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Behind the scenes}

\begin{itemize}
\item polymorphic macros and interaction with type inference
\item reify defeats erasure
\item ultra-cake pattern in the new Scala reflection
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Future work}

\begin{itemize}
\item Marriage of macros and type inference
\item Better reify
\item Macro types
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Thanks!}

\centering
\text{\color{linkblue}\href{mailto:eugene.burmako@epfl.ch}{eugene.burmako@epfl.ch}}

\end{frame}

\end{document}
